\documentclass[11pt]{article}
\usepackage{amsmath, amssymb}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{parskip}

\geometry{margin=1.1in}

\title{\textbf{Hog Population Simulation: Model Description}}
\date{}

\begin{document}
\maketitle

\tableofcontents
\bigskip

% ─────────────────────────────────────────────────────────────────────────────
\section{World Representation}

The simulation world is a discrete grid of $(S_x + 1) \times (S_y + 1)$ nodes, with
$S_x = S_y = 100$.  Every scalar field in the model---agriculture, avoidance,
hog density---is stored at these nodes and indexed by integer column $c$ and
row $r$, with $0 \le c \le S_x$ and $0 \le r \le S_y$.

Each hog carries a continuous position $(x, y) \in [0, S_x] \times [0, S_y]$;
the nearest-integer rounding $c = \mathrm{round}(x)$, $r = \mathrm{round}(y)$
is used whenever a grid-node value is looked up.

% ─────────────────────────────────────────────────────────────────────────────
\section{Static Fields}

\subsection{Agriculture, $A(c,r)$}

Agriculture encodes local food availability.  It is generated once at
initialisation as a superposition of $W = 8$ random cosine waves:

\[
  A_{\mathrm{raw}}(c,r)
  \;=\; \sum_{w=1}^{W}
        \cos\!\bigl(c\,f_x^{(w)} + r\,f_y^{(w)} + \varphi^{(w)}\bigr),
\]

where each spatial frequency $f_x^{(w)}, f_y^{(w)} \sim \mathrm{Uniform}(-0.09,\,0.09)$
and each phase $\varphi^{(w)} \sim \mathrm{Uniform}(0, 2\pi)$.  The raw field is
then min-max normalised:

\[
  A(c,r) \;=\; \frac{A_{\mathrm{raw}}(c,r) - \min A_{\mathrm{raw}}}
                    {\max A_{\mathrm{raw}} - \min A_{\mathrm{raw}}}
  \;\in\; [0,\,1].
\]

The superposition of low-frequency cosines produces smooth, blob-like spatial
variation without sharp discontinuities.

\subsection{Agriculture Gradient}

Because hogs steer toward increasing food, the spatial gradient of $A$ is
precomputed once via central differences:

\[
  \nabla_c A(c,r)
    = A(\min(c{+}1, S_x),\, r) - A(\max(c{-}1,0),\, r),
\]
\[
  \nabla_r A(c,r)
    = A(c,\, \min(r{+}1, S_y)) - A(c,\, \max(r{-}1,0)).
\]

Because $A$ is static, this computation is performed only at startup.

\subsection{Avoidance Field, $V(c,r)$}

An avoidance field $V(c,r) \in [0,1]$ encodes locations that hogs should avoid
(e.g.\ human infrastructure, hunting zones).  Unlike agriculture, avoidance is
not used via its gradient.  Instead, hogs sense danger through a direct
radius search described in Section~\ref{sec:avoid_dir}, which produces a
repulsive flee direction that is nonzero at any distance up to $R_{\mathrm{avoid}}$
from a hazard.  A gradient-based approach would produce a nonzero signal only
at the immediate boundary between zero and nonzero cells, making it effectively
inactive for a field that is mostly zero.

% ─────────────────────────────────────────────────────────────────────────────
\section{Hog State}

Each hog $i$ is described by:

\begin{itemize}
  \item \textbf{Position} $(x_i, y_i) \in [0, S_x] \times [0, S_y]$, continuous.
  \item \textbf{Fatness} $\phi_i \in [0, 1]$, a proxy for energy reserves.
  \item \textbf{Direction} $(\hat{d}_{x,i},\, \hat{d}_{y,i})$, a unit vector
        computed each time-step before any hog is moved.
\end{itemize}

At initialisation, fatness is drawn uniformly from $[0.35,\, 0.70]$ and hogs
are placed at grid nodes drawn uniformly at random.

% ─────────────────────────────────────────────────────────────────────────────
\section{Direction Calculation}

The direction of hog $i$ is the normalised weighted sum of four deterministic
components---pack cohesion, food-seeking, threat avoidance, and crowd
dispersal---plus a stochastic wander term.

\subsection{Pack Cohesion}

Let $\mathcal{N}_i$ be the set of the $N_{\mathrm{pack}} = 5$ nearest
neighbours of hog $i$.  The cohesion direction is

\[
  \mathbf{g}_i = \frac{\tilde{\mathbf{g}}_i}{\|\tilde{\mathbf{g}}_i\|},
  \qquad
  \tilde{\mathbf{g}}_i
    = \sum_{j \in \mathcal{N}_i} \frac{\mathbf{p}_j - \mathbf{p}_i}
                                      {\|\mathbf{p}_j - \mathbf{p}_i\|},
\]

where $\mathbf{p}_k = (x_k, y_k)$.  Normalising each pairwise displacement
before summing gives each neighbour equal angular vote regardless of distance,
so a single distant outlier does not dominate.

\subsection{Food-Seeking}

The food-seeking direction is the normalised agriculture gradient at the hog's
current node:

\[
  \mathbf{a}_i
    = \frac{\nabla A(c_i, r_i)}{\|\nabla A(c_i, r_i)\|}.
\]

Hogs that are hungrier (lower $\phi_i$) weight this component more heavily;
see Section~\ref{sec:weights}.

\subsection{Threat Avoidance}
\label{sec:avoid_dir}

The flee direction is computed by a radius search over all grid nodes within
$R_{\mathrm{avoid}}$ cells that carry nonzero avoidance.  Each such node
contributes a unit vector pointing away from it, weighted by $V / \mathrm{dist}$
so that closer and stronger hazards exert greater influence:

\[
  \tilde{\mathbf{f}}_i
    = \sum_{\substack{(n_c,\, n_r):\\ \delta \;\leq\; R_{\mathrm{avoid}} \\[2pt] V(n_c,n_r) > 0}}
      \frac{(c_i - n_c,\; r_i - n_r)}{\delta}
      \cdot \frac{V(n_c, n_r)}{\delta},
  \qquad
  \delta = \bigl\|(c_i - n_c,\; r_i - n_r)\bigr\|,
\]

\[
  \mathbf{f}_i = \frac{\tilde{\mathbf{f}}_i}{\|\tilde{\mathbf{f}}_i\|}.
\]

When no hazard exists within $R_{\mathrm{avoid}}$, $\tilde{\mathbf{f}}_i = \mathbf{0}$
and the term contributes nothing to the direction sum.

\subsection{Crowd Dispersal}

Let $\mathcal{C}_i$ be the set of all hogs within radius $R_{\mathrm{crowd}} = 8$
grid cells \emph{excluding} the $N_{\mathrm{pack}}$ nearest neighbours.  The
dispersal direction pushes hog $i$ away from the crowd centroid:

\[
  \tilde{\mathbf{c}}_i
    = -\sum_{j \in \mathcal{C}_i}
        \frac{\mathbf{p}_j - \mathbf{p}_i}{\|\mathbf{p}_j - \mathbf{p}_i\|},
  \qquad
  \mathbf{c}_i = \frac{\tilde{\mathbf{c}}_i}{\|\tilde{\mathbf{c}}_i\|}.
\]

The severity of crowd pressure is

\[
  \sigma_i = \min\!\left(\frac{|\mathcal{C}_i|}{N_{\mathrm{pack}} + 1},\; 3\right),
\]

which grows with crowd size and saturates at 3.  Division by
$N_{\mathrm{pack}} + 1$ sets the scale: a crowd equal in number to the pack
produces $\sigma = 1$, a crowd twice as large produces $\sigma = 2$, and so on.

\subsection{Wander}

A stochastic impulse $\mathbf{w}_i = (\cos\theta,\,\sin\theta)$ with
$\theta \sim \mathrm{Uniform}(0, 2\pi)$ is added at each step.  Its weight
is drawn as:

\[
  w_{\mathrm{wander}} =
  \begin{cases}
    \mathrm{Uniform}(W_{\mathrm{hi,min}},\, W_{\mathrm{hi,max}})
      & \text{with probability } p_{\mathrm{wander}} = 0.05,\\[4pt]
    \mathrm{Uniform}(0,\, W_{\mathrm{lo}})
      & \text{otherwise.}
  \end{cases}
\]

In the triggered case the weight can reach $W_{\mathrm{hi,max}} = 6$, which
can overpower all deterministic components and cause sudden re-orientation.
In the quiet case the weight is at most $W_{\mathrm{lo}} = 0.15$, acting as
negligible noise.

\subsection{Weighted Combination}
\label{sec:weights}

The raw direction vector is

\begin{equation}
\label{eq:direction}
  \tilde{\mathbf{d}}_i
    = W_g\,\mathbf{g}_i
    + W_a(\phi_i)\,\mathbf{a}_i
    + W_v\,\mathbf{f}_i
    + W_c\,\sigma_i\,\mathbf{c}_i
    + w_{\mathrm{wander}}\,\mathbf{w}_i,
\end{equation}

with the hunger-scaled food weight

\[
  W_a(\phi_i) = W_a\,(2 - \phi_i).
\]

This doubles the food weight when $\phi_i = 0$ (starving) and reduces it to
$W_a$ when $\phi_i = 1$ (fully fed).  Because $\mathbf{f}_i = \mathbf{0}$ when
no danger is within $R_{\mathrm{avoid}}$, the avoidance term is naturally
inactive in safe territory without requiring an additional multiplier.  The
final unit direction is

\[
  (\hat{d}_{x,i}, \hat{d}_{y,i}) = \frac{\tilde{\mathbf{d}}_i}{\|\tilde{\mathbf{d}}_i\|}.
\]

% ─────────────────────────────────────────────────────────────────────────────
\section{Movement}

All hog directions are computed from the same position snapshot before any hog
is moved (synchronous update).  Hog $i$ then advances by

\[
  \mathbf{p}_i \;\leftarrow\;
    \mathrm{clamp}\!\Bigl(
      \mathbf{p}_i
      + v(\phi_i)\,\Delta t \cdot (\hat{d}_{x,i}, \hat{d}_{y,i}),
      \;\mathbf{0},\;(S_x, S_y)
    \Bigr),
\]

where the speed is

\[
  v(\phi_i) = v_0\,\bigl(1 - \tfrac{1}{2}\,\phi_i\bigr),
  \qquad v_0 = 5 \text{ grid cells day}^{-1}.
\]

Fatter hogs move at half the base speed.

% ─────────────────────────────────────────────────────────────────────────────
\section{Fatness Dynamics}

At each time-step, hog $i$'s fatness changes as

\[
  \phi_i \;\leftarrow\;
  \mathrm{clamp}\!\Bigl(
    \phi_i
    + \bigl(A(c_i,r_i) - 0.5\bigr)\cdot 2\,U\,\Delta t,
    \;0,\;1
  \Bigr),
\]

where $U = 0.08$ day$^{-1}$ is the maximum fat-change rate.  Subtracting 0.5
recenters the agriculture value so that $A = 0.5$ is neutral; multiplying by 2
stretches the range to $[-1, 1]$; multiplying by $U$ scales to
$[-0.08,\, 0.08]$ per day.  At $A = 1$ a hog gains $U$ fat per day; at
$A = 0$ it loses $U$ per day.  Fatness is hard-capped at 1.  If $\phi_i$
reaches 0 the hog dies of starvation.

% ─────────────────────────────────────────────────────────────────────────────
\section{Demography}

\subsection{Natural Mortality}

Each hog faces a constant background death hazard $\mu$ per day:

\[
  P(\text{death in } \Delta t) = \mu\,\Delta t,
  \qquad
  \mu = \frac{\lambda}{R^*},
\]

where $\lambda$ is the birth rate and $R^* = 1.05$ is the target maximum
birth-to-death ratio.  The net per-capita growth rate under ideal conditions
(starvation absent) is

\[
  \lambda - \mu = \lambda\!\left(1 - \frac{1}{R^*}\right)
               = \frac{1}{12} \cdot \frac{0.05}{1.05}
               \approx 0.004 \text{ day}^{-1}.
\]

Under poor conditions starvation mortality adds to $\mu$ and the population
contracts.

\subsection{Reproduction}

Each surviving hog reproduces at rate $\lambda = \tfrac{1}{12}$ day$^{-1}$,
independent of local food conditions.  A birth occurs within $\Delta t$ with
probability $\lambda\,\Delta t$.  The offspring is placed at the parent's
current node with initial fatness

\[
  \phi_{\mathrm{child}} = \frac{\phi_{\mathrm{parent}} + 0.5}{2},
\]

regressing toward the population mean.  Reproduction is suppressed when the
total population reaches the hard cap $N_{\mathrm{max}} = 600$.

% ─────────────────────────────────────────────────────────────────────────────
\section{Time Discretisation}

Simulated time is measured in days.  The real-time day length is
$T_{\mathrm{day}} = 2000\,\text{ms}$.  Each frame supplies an elapsed
wall-clock time $\Delta t_{\mathrm{wall}}$ (ms), giving

\[
  \Delta t = \frac{\Delta t_{\mathrm{wall}}}{T_{\mathrm{day}}}.
\]

All per-day rates are multiplied by $\Delta t$ before application, producing a
continuous-time Euler integration that is independent of frame rate.

% ─────────────────────────────────────────────────────────────────────────────
\section{Summary of Constants}

\begin{center}
\begin{tabular}{llll}
\toprule
Symbol & Value & Unit & Description \\
\midrule
$S_x, S_y$               & 100               & ---              & Grid dimensions \\
$N_{\mathrm{pop,0}}$     & 150               & hogs             & Initial population \\
$N_{\mathrm{max}}$       & 600               & hogs             & Population hard cap \\
$T_{\mathrm{day}}$       & 2000              & ms               & Real-time day length \\
$v_0$                    & 5                 & cells day$^{-1}$ & Base movement speed \\
$N_{\mathrm{pack}}$      & 5                 & hogs             & Pack size \\
$R_{\mathrm{crowd}}$     & 8                 & cells            & Crowd dispersal radius \\
$R_{\mathrm{avoid}}$     & 15                & cells            & Avoidance sensing radius \\
$W_g$                    & 1.0               & ---              & Pack cohesion weight \\
$W_a$                    & 1.5               & ---              & Agriculture weight (base) \\
$W_v$                    & 2.0               & ---              & Avoidance weight \\
$W_c$                    & 2.5               & ---              & Crowd dispersal weight \\
$p_{\mathrm{wander}}$    & 0.05              & ---              & Wander trigger probability \\
$W_{\mathrm{lo}}$        & 0.15              & ---              & Quiet wander max weight \\
$W_{\mathrm{hi,min}}$    & 2.0               & ---              & Triggered wander min weight \\
$W_{\mathrm{hi,max}}$    & 6.0               & ---              & Triggered wander max weight \\
$U$                      & 0.08              & day$^{-1}$       & Max fat change rate \\
$\lambda$                & $1/12$            & day$^{-1}$       & Birth rate \\
$R^*$                    & 1.05              & ---              & Max birth-to-death ratio \\
$\mu$                    & $\lambda / R^*$   & day$^{-1}$       & Natural death rate \\
\bottomrule
\end{tabular}
\end{center}

% ─────────────────────────────────────────────────────────────────────────────
\section{Implementation Notes}

\subsection{Synchronous Update Order}

Within each time-step the simulation proceeds in four sequential phases:
(1) all directions are computed from the current position snapshot;
(2) all hogs are moved;
(3) vital events (starvation, natural death) are evaluated;
(4) reproduction is evaluated.
Separating direction calculation from movement ensures that no hog benefits
from the updated positions of hogs that moved earlier in the same step.

\subsection{Spatial Index}

A two-dimensional array indexed by $(c, r)$ stores references to all hogs
whose rounded grid position is that node.  This index is maintained
incrementally as hogs move.

\subsection{Neighbour Search}

Direction calculation requires the $N_{\mathrm{pack}}$ nearest neighbours and
the crowd set within $R_{\mathrm{crowd}}$.  Both are obtained from a single
distance-sorted pass over all hogs.  The crowd loop terminates early once
distances exceed $R_{\mathrm{crowd}}$.  The avoidance radius search iterates
over grid nodes in the bounding box $[c - R_{\mathrm{avoid}},\,
c + R_{\mathrm{avoid}}] \times [r - R_{\mathrm{avoid}},\, r +
R_{\mathrm{avoid}}]$, skipping nodes where $V = 0$.

\subsection{Avoidance Placement}

Avoidance splotches are painted onto $V$ interactively.  A circular splotch of
radius $R_{\mathrm{splotch}} = 8$ cells is placed at a clicked grid location,
with a linear falloff from the specified peak value $v_0 \in (0, 1]$ at the
centre to zero at the perimeter:

\[
  V(n_c, n_r) \;\leftarrow\;
  \min\!\Bigl(1,\; V(n_c, n_r)
    + v_0\,\Bigl(1 - \frac{\|(n_c - c_0,\, n_r - r_0)\|}{R_{\mathrm{splotch}}}\Bigr)\Bigr),
\]

for all nodes within $R_{\mathrm{splotch}}$ of the click centre $(c_0, r_0)$.
Values accumulate additively and are clamped to $[0, 1]$.

\end{document}